---
title: "Final Project"
author: "Shriya Jain"
output: word_document
---
# LOADING THE PACKAGES 
```{r}
# Loading all the required packages 
packages <- c(
  "tidyverse",
  "janitor",
  "NHANES",
  "naniar",
  "ResourceSelection",
  "pROC",
  "car",
  "broom",
  "forcats",
  "patchwork"
)

installed <- rownames(installed.packages())
for (p in packages) {
  if (!(p %in% installed)) install.packages(p)
}

library(tidyverse)
library(janitor)
library(NHANES)
library(naniar)
library(ResourceSelection)
library(pROC)
library(car)
library(broom)
library(forcats)
library(patchwork)
library(GGally)
```

# LOADING THE DATA
```{r}
# Loading the NHANES data
data("NHANES")

# Cleaning the column names
nhanes <- NHANES %>%
  clean_names()

# Checking the dimensions of the data
dim(nhanes)
```

# KEEPING ONE ROW PER PERSON
```{r}
# Keeping the distinct ids 
nhanes_unique <- nhanes %>%
  distinct(id, .keep_all = TRUE)

# Checking the dimensions after removing duplicate ids
dim(nhanes_unique)
```

# SELECTING THE VARIABLES OF INTEREST 
```{r}
# Selecting the variables from the NHANES data 
dat <- nhanes_unique %>%
  select(
    id,
    survey_yr,
    age,
    gender,
    race1,
    education,
    hh_income,
    poverty,
    bmi,
    bmi_who,
    bp_sys_ave,
    bp_dia_ave,
    tot_chol,
    diabetes,
    smoke_now,
    alcohol12plus_yr,
    sleep_hrs_night,
    phys_active
  )
```

# FILTERING THE DATA BASED ON AGE AND BMI CUTOFFS
```{r}
dat <- dat %>%
  filter(age >= 18) %>%
  mutate(
    # Filtering for hypertension with systolic >= 140 or diastolic >= 90
    hypertension = case_when(
      bp_sys_ave >= 140 | bp_dia_ave >= 90 ~ 1,
      !is.na(bp_sys_ave) & !is.na(bp_dia_ave) ~ 0,
      TRUE ~ NA_real_
    ),

    # Filtering based on the standard adult cutoffs for BMI
    bmi_category = case_when(
      bmi < 18.5 ~ "Underweight",
      bmi >= 18.5 & bmi < 25 ~ "Normal",
      bmi >= 25 & bmi < 30 ~ "Overweight",
      bmi >= 30 ~ "Obese",
      TRUE ~ NA_character_
    )
  )
```

# CONVERTING THE VARIABLES INTO FACTORS 
```{r}
dat <- dat %>%
  mutate(
    survey_yr = factor(survey_yr),
    gender = factor(gender),
    race1 = factor(race1),
    education = factor(education),
    hh_income = factor(hh_income),
    diabetes = factor(diabetes),
    smoke_now = factor(smoke_now),
    alcohol12plus_yr = factor(alcohol12plus_yr),
    phys_active = factor(phys_active),
    hypertension = factor(hypertension, levels = c(0, 1), labels = c("No", "Yes")),
    bmi_category = factor(
      bmi_category,
      levels = c("Normal", "Underweight", "Overweight", "Obese")
    )
  )

# Checking the cleaned dataset
summary(dat)
```

# CHECKING FOR THE MISSING DATA
```{r}
# Checking the missing values by variable 
colSums(is.na(dat))
```

# IMPUTING FOR THE MISSING DATA
```{r}
dat_imp <- dat

# Creating a function to calculate mode 
get_mode <- function(x) {
  ux <- unique(x[!is.na(x)])
  ux[which.max(tabulate(match(x, ux)))]
}

# Formulating mean imputation for numeric variables
numeric_vars <- names(dat_imp)[sapply(dat_imp, is.numeric)]
for (v in numeric_vars) {
  dat_imp[[v]][is.na(dat_imp[[v]])] <- mean(dat_imp[[v]], na.rm = TRUE)
}

# Formulating mode imputation for categorical variables
factor_vars <- names(dat_imp)[sapply(dat_imp, is.factor)]
for (v in factor_vars) {
  mode_val <- get_mode(dat_imp[[v]])
  dat_imp[[v]][is.na(dat_imp[[v]])] <- mode_val
}

# Confirming the results after imputation
colSums(is.na(dat_imp))
```

# EXPLORATORY DATA ANALYSIS
```{r}
# Creating age groups for subgroup comparisons
dat_imp <- dat_imp %>%
  mutate(
    age_group = cut(
      age,
      breaks = c(18, 30, 45, 60, Inf),
      labels = c("18-29", "30-44", "45-59", "60+"),
      right = FALSE
    )
  )

# Creating consistent color for hypertension
hyp_cols <- c(
  "No" = "#F8766D",
  "Yes" = "#00BFC4"
)

# Creating consistent color for sex
sex_cols <- c(
  "female" = "#E69F00",
  "male"   = "#56B4E9"
)

# Creating consistent color for age groups 
age_cols <- c(
  "18-29" = "#4E79A7",
  "30-44" = "#59A14F",
  "45-59" = "#F28E2B",
  "60+"   = "#E15759"
)

# Creating consistent color for BMI
bmi_cat_cols <- c(
  "Underweight" = "#8DD3C7",
  "Normal"      = "#80B1D3",
  "Overweight"  = "#FDB462",
  "Obese"       = "#FB8072"
)

# Creating consistent color for race
race_cols <- c(
  "Mexican"    = "#66C2A5",
  "Hispanic"      = "#FC8D62",
  "White"  = "#8DA0CB",
  "Black"  = "#E78AC3",
  "Other"          = "#A6D854"
)

common_theme <- theme_minimal(base_size = 14)

# DISTRIBUTION PLOTS
# Distribution of BMI
dist_plot_bmi <- ggplot(dat_imp, aes(x = bmi)) +
  geom_histogram(binwidth = 2, fill = "#4E79A7", color = "black") +
  labs(
    title = "BMI Distribution",
    x = "Body Mass Index (BMI)",
    y = "Count"
  ) +
  common_theme

# Distribution of Age
dist_plot_age <- ggplot(dat_imp, aes(x = age)) +
  geom_histogram(binwidth = 2, fill = "#F28E2B", color = "black") +
  labs(
    title = "Age Distribution",
    x = "Age",
    y = "Count"
  ) +
  common_theme

# Distribution of Hypertension
dist_plot_hyp <- ggplot(dat_imp, aes(x = hypertension, fill = hypertension)) +
  geom_bar() +
  scale_fill_manual(values = hyp_cols) +
  labs(
    title = "Hypertension Distribution",
    x = "Hypertension Status",
    y = "Count"
  ) +
  common_theme +
  theme(legend.position = "none")

# PLOTS FOR HYPERTENSION AND AGE, SEX, RACE
# Hypertension proportion by age group
plot_hyp_age <- ggplot(dat_imp, aes(x = age_group, fill = hypertension)) +
  geom_bar(position = "fill") +
  scale_fill_manual(values = hyp_cols) +
  labs(
    title = "Hypertension Proportion by Age Group",
    x = "Age Group",
    y = "Proportion",
    fill = "Hypertension"
  ) +
  common_theme

# Hypertension proportion by sex
plot_hyp_sex <- ggplot(dat_imp, aes(x = gender, fill = hypertension)) +
  geom_bar(position = "fill") +
  scale_fill_manual(values = hyp_cols) +
  labs(
    title = "Hypertension Proportion by Sex",
    x = "Sex",
    y = "Proportion",
    fill = "Hypertension"
  ) +
  common_theme

# Hypertension proportion by race
plot_hyp_race <- ggplot(dat_imp, aes(x = race1, fill = hypertension)) +
  geom_bar(position = "fill") +
  scale_fill_manual(values = hyp_cols) +
  labs(
    title = "Hypertension Proportion by Race",
    x = "Race",
    y = "Proportion",
    fill = "Hypertension"
  ) +
  common_theme +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# RELATIONSHIP PLOTS
# Plot for capturing BMI and hypertension 
plot_bmi_hyp <- ggplot(dat_imp, aes(x = hypertension, y = bmi, fill = hypertension)) +
  geom_boxplot() +
  scale_fill_manual(values = hyp_cols) +
  labs(
    title = "BMI by Hypertension Status",
    x = "Hypertension Status",
    y = "BMI"
  ) +
  common_theme +
  theme(legend.position = "none")

# Plot for capturing systolic blood pressure and BMI
plot_sys_bmi <- ggplot(dat_imp, aes(x = bmi_category, y = bp_sys_ave, fill = bmi_category)) +
  geom_boxplot() +
  scale_fill_manual(values = bmi_cat_cols) +
  labs(
    title = "Systolic Blood Pressure by BMI Category",
    x = "BMI Category",
    y = "Average Systolic Blood Pressure"
  ) +
  common_theme +
  theme(legend.position = "none")

# Plot for capturing BMI and age group
plot_bmi_age <- ggplot(dat_imp, aes(x = age_group, y = bmi, fill = age_group)) +
  geom_boxplot() +
  scale_fill_manual(values = age_cols) +
  labs(
    title = "BMI by Age Group",
    x = "Age Group",
    y = "BMI"
  ) +
  common_theme +
  theme(legend.position = "none")

# Plot for capturing systolic blood pressure and age group
plot_sys_age <- ggplot(dat_imp, aes(x = age_group, y = bp_sys_ave, fill = age_group)) +
  geom_boxplot() +
  scale_fill_manual(values = age_cols) +
  labs(
    title = "Systolic Blood Pressure by Age Group",
    x = "Age Group",
    y = "Average Systolic Blood Pressure"
  ) +
  common_theme +
  theme(legend.position = "none")

# Plot for capturing BMI and sex
plot_bmi_sex <- ggplot(dat_imp, aes(x = gender, y = bmi, fill = gender)) +
  geom_boxplot() +
  scale_fill_manual(values = sex_cols) +
  labs(
    title = "BMI by Sex",
    x = "Sex",
    y = "BMI"
  ) +
  common_theme +
  theme(legend.position = "none")

# Plot for capturing systolic blood pressure and sex
plot_sys_sex <- ggplot(dat_imp, aes(x = gender, y = bp_sys_ave, fill = gender)) +
  geom_boxplot() +
  scale_fill_manual(values = sex_cols) +
  labs(
    title = "Systolic Blood Pressure by Sex",
    x = "Sex",
    y = "Average Systolic Blood Pressure"
  ) +
  common_theme +
  theme(legend.position = "none")

# Plot for capturing BMI and race
plot_bmi_race <- ggplot(dat_imp, aes(x = race1, y = bmi, fill = race1)) +
  geom_boxplot() +
  scale_fill_manual(values = race_cols) +
  labs(
    title = "BMI by Race",
    x = "Race",
    y = "BMI"
  ) +
  common_theme +
  theme(
    legend.position = "none",
    axis.text.x = element_text(angle = 45, hjust = 1)
  )

# Plot for capturing systolic blood pressure and race
plot_sys_race <- ggplot(dat_imp, aes(x = race1, y = bp_sys_ave, fill = race1)) +
  geom_boxplot() +
  scale_fill_manual(values = race_cols) +
  labs(
    title = "Systolic Blood Pressure by Race",
    x = "Race",
    y = "Average Systolic Blood Pressure"
  ) +
  common_theme +
  theme(
    legend.position = "none",
    axis.text.x = element_text(angle = 45, hjust = 1)
  )


# PRINTING THE PLOTS
# Printing the distribution plots
dist_plot_bmi
dist_plot_age
dist_plot_hyp
plot_hyp_age
plot_hyp_sex
plot_hyp_race

# Printing the relationship plots 
plot_bmi_hyp
plot_sys_bmi
plot_bmi_age
plot_sys_age
plot_bmi_sex
plot_sys_sex
plot_bmi_race
plot_sys_race
```

```{r}
# Correlation Matrix
dat_imp %>%
  select(bmi, age, bp_sys_ave, bp_dia_ave, tot_chol) %>%
  ggpairs(title = "Correlation Matrix of Continuous Variables")
```

# ADDING DESCRIPTIVE SUMMARIES 
```{r}
summary(dat_imp)
```

# STRATISFIED SUMMARIES
```{r}
# Summary for BMI and hypertension
dat_imp %>%
  group_by(hypertension) %>%
  summarise(
    mean_bmi = mean(bmi, na.rm = TRUE),
    sd_bmi = sd(bmi, na.rm = TRUE)
  )

# Summary for age and hypertension
dat_imp %>%
  group_by(hypertension) %>%
  summarise(
    mean_age = mean(age, na.rm = TRUE),
    sd_age = sd(age, na.rm = TRUE)
  )

# Summary for BP and hypertension
dat_imp %>%
  group_by(hypertension) %>%
  summarise(
    mean_sys = mean(bp_sys_ave, na.rm = TRUE),
    mean_dia = mean(bp_dia_ave, na.rm = TRUE)
  )
```

# CATEGORICAL SUMMARIES
```{r}
# Summary for distribution of sex
prop.table(table(dat_imp$gender))

# Summary for distribution of race
prop.table(table(dat_imp$race1))

# Summary for distribution of hypertension prevelance
prop.table(table(dat_imp$hypertension))
```

# FITTING THE LOGISTIC REGRESSION MODELS 
```{r}
# Creating model 1 for baseline 
model1 <- glm(hypertension ~ 1, data = dat_imp, family = binomial)

# Creating model 2 with BMI as a continuous predictor
model2 <- glm(hypertension ~ bmi + age + gender + race1,
              data = dat_imp,
              family = binomial)

# Creating model 3 with BMI as a categorical predictor
model3 <- glm(hypertension ~ bmi_category + age + gender + race1,
              data = dat_imp,
              family = binomial)
```

```{r}
# Getting a summary for all the models 
summary(model1)
summary(model2)
summary(model3)
```

# COMPUTING THE ODDS RATIO AND 95% CI
```{r}
# Creating a function to extract odds ratios
or_table <- function(model) {
  tidy(model, conf.int = TRUE, exponentiate = TRUE) %>%
    rename(
      odds_ratio = estimate,
      conf_low = conf.low,
      conf_high = conf.high
    )
}

# Creating a table to capture the odds ratios for each model
or_model1 <- or_table(model1)
or_model2 <- or_table(model2)
or_model3 <- or_table(model3)

# Printing the tables 
or_model1
or_model2
or_model3
```

# COMPARISON BETWEEN DIFFERENT MODELS 
```{r}
# Comparing baseline model to continuous BMI model using likelihood ratio test
anova(model1, model2, test = "LRT")

# Comparing baseline model to categorical BMI model using likelihood ratio test
anova(model1, model3, test = "LRT")

# Comparing AIC and BIC across all models
AIC(model1, model2, model3)
BIC(model1, model2, model3)
```

# PREPARING NUMERIC HYPERTENSION OUTCOME FOR DIAGNOSTICS 
```{r}
# Creating numeric hypertension variable for diagnostics
dat_imp <- dat_imp %>%
  mutate(htn_num = ifelse(hypertension == "Yes", 1, 0))
```

# COMPUTING THE HOSMER LEMESHOW TEST 
```{r}
# Computing Hosmer-Lemeshow goodness-of-fit test for model 2 and model 3
hoslem.test(dat_imp$htn_num, fitted(model2), g = 10)
hoslem.test(dat_imp$htn_num, fitted(model3), g = 10)
```

# FORMULATING THE ROC CURVES AND AUC 
```{r}
# Creating ROC objects for each model
roc1 <- roc(dat_imp$htn_num, fitted(model1))
roc2 <- roc(dat_imp$htn_num, fitted(model2))
roc3 <- roc(dat_imp$htn_num, fitted(model3))

# Printing AUC values
auc(roc1)
auc(roc2)
auc(roc3)

# Plotting ROC curves
plot(roc1, main = "ROC Curve Comparison for Logistic Regression Models", col = "blue")
plot(roc2, add = TRUE, col = "red")
plot(roc3, add = TRUE, col = "darkgreen")

legend(
  "bottomright",
  legend = c(
    paste("Model 1 AUC =", round(auc(roc1), 3)),
    paste("Model 2 AUC =", round(auc(roc2), 3)),
    paste("Model 3 AUC =", round(auc(roc3), 3))
  ),
  col = c("blue", "red", "darkgreen"),
  lty = 1,
  cex = 0.8
)
```
# CHECKING FOR MULTICOLLINEARITY 
```{r}
# Checking for the variance influence factors for the models 
vif(model2)
vif(model3)
```

# CHECKING THE LINEARITY FOR CONTINUOUS PREDICTORS 
```{r}
# Formulating the predicted probabilties from model 2
dat_imp$pred_prob <- fitted(model2)

pred_race <- dat_imp %>%
  group_by(race1) %>%
  summarise(pred_prob = mean(pred_prob, na.rm = TRUE))

# Plot for predictive probability of hypertension and BMI
ggplot(dat_imp, aes(x = bmi, y = pred_prob)) +
  geom_point(alpha = 0.4, color = "steelblue") +
  geom_smooth(se = TRUE, color = "red") +
  labs(
    title = "Predicted Probability of Hypertension and BMI",
    x = "BMI",
    y = "Predicted Probability"
  ) +
  theme_minimal()

# Plot for predictive probability of hypertension by BMI and age
ggplot(dat_imp, aes(x = age, y = pred_prob)) +
  geom_point(alpha = 0.4, color = "darkorange") +
  geom_smooth(se = TRUE, color = "red") +
  labs(
    title = "Predicted Probability of Hypertension by BMI and Age",
    x = "Age",
    y = "Predicted Probability"
  ) +
  theme_minimal()

# Plot for predictive probability of hypertension by BMI and sex
ggplot(dat_imp, aes(x = bmi, y = pred_prob, color = gender)) +
  geom_point(alpha = 0.4) +
  geom_smooth(se = FALSE) +
  labs(
    title = "Predicted Probability of Hypertension by BMI and Sex",
    x = "BMI",
    y = "Predicted Probability",
    color = "Sex"
  ) +
  theme_minimal()


# Plot for predictive probability of hypertension by BMI and race
ggplot(dat_imp, aes(x = bmi, y = pred_prob, color = race1)) +
  geom_point(alpha = 0.15) +
  geom_smooth(se = FALSE, linewidth = 1) +
  labs(
    title = "Predicted Probability of Hypertension by BMI and Race",
    x = "BMI",
    y = "Predicted Probability",
    color = "Race"
  ) +
  theme_minimal(base_size = 14)
```

# FORMULATING THE FOREST PLOT FOR ODDS RATIO
```{r}
# Extracting ORs from model2
forest_data <- tidy(model2, exponentiate = TRUE, conf.int = TRUE) %>%
  filter(term != "(Intercept)")

# Cleaning the labels
forest_data$term <- factor(forest_data$term, levels = rev(forest_data$term))

# Creating a forest plot for the odds ratios
ggplot(forest_data, aes(x = term, y = estimate)) +
  geom_point(color = "darkblue", size = 2) +
  geom_errorbar(aes(ymin = conf.low, ymax = conf.high), width = 0.2) +
  geom_hline(yintercept = 1, linetype = "dashed", color = "red") +
  coord_flip() +
  labs(
    title = "Odds Ratios for Hypertension",
    x = "Predictors",
    y = "Odds Ratio (95% CI)"
  ) +
  theme_minimal()
```

